---
title: "split-apply-combine"
output:
  html_document:
    toc: yes
    number_sections: yes
  html_notebook:
    toc: yes
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0;
}
</style>

Earl F Glynn | KC R Users Meetup | 2019-07-20

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
time.1 <- Sys.time()
```

```{r, comment = NA}
library(tidyverse)
library(kableExtra) # kable_styling
```

Helper function

```{r}
Show <- function(data, caption="", bigMark="", ...)
{
  data                                          %>%
  kable("html",
        caption=caption,
        format.args = list(big.mark=bigMark))   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left", full_width = FALSE, ...)
}
```

# Background

* torvaney's RStudio posting, [Apply function to dataframe in chunks](https://community.rstudio.com/t/apply-function-to-dataframe-in-chunks/9067).

* [Split-Apply-Combine: My search for a replacement for 'group_by + do'](https://coolbutuseless.bitbucket.io/2018/03/03/split-apply-combine-my-search-for-a-replacement-for-group_by---do/).

# Data to Explore:  `iris` data.frame

Let's use the `iris` data.frame that has three groups of iris data.

```{r, comment = NA}
str(iris)
```

# dplyr summarize function

```{r}
iris                 %>%
  group_by(Species)  %>%
  summarize(                   # Also spelled `summarise`
             n = n(), 
             meanSepalLength = mean(Sepal.Length, na.rm = TRUE),
             meanSepalWidth  = mean(Sepal.Width,  na.rm = TRUE),
             meanPetalLength = mean(Petal.Length, na.rm = TRUE),
             meanPetalWidth  = mean(Petal.Width,  na.rm = TRUE)
           )        %>%
  ungroup()         %>%
  Show()
```

Functions used in `summarize` have to return a scalar value, not a vector or other object.

What if a special function is needed to summarize/manipulate the data by groups?

What if we want various quantiles of one of the variables for all groups and only want to call the `quantile` function once?

How can we process group data with a result as another tibble?

# Split-Apply-Combine - Pre-historic R method

Pre-tidyverse approach

## Split into list of groups

```{r, comment = NA}
splits <- split(iris, iris$Species)
length(splits)
```

```{r, comment = NA}
dim(splits[[1]])
```

```{r}
splits[[1]] %>% head(2)  %>% Show()
```

## Write function that processes one of the groups

Let's assume we want several stats for one of the columns 

```{r}
x <- splits[[1]]
```

Here `x` is a data.frame with all data for one group.

Example quantiles

```{r, comment = NA}
quantile(x[, "Sepal.Length"], 
         c(0.00, 0.05, 0.25, 0.50, 0.75, 0.95, 1.00))
```

Let's write a function that processes a data.frame `x` with the specific ``group` column`columnName`

```{r}
getStats <- function(x, columnName)
{
  xValue <- x[, columnName]
  
  Q <- quantile(xValue, 
                c(0.00, 0.05, 0.25, 0.50,
                  0.75, 0.95, 1.00),
                na.rm = TRUE)
  
  tibble(
    Species = x$Species[1],
    n     = nrow(x),
    
    mean  = mean(xValue, na.rm = TRUE),
    stdev = sd(xValue,   na.rm = TRUE),
    
    min  = Q[1],
    Q005 = Q[2],
    Q025 = Q[3],
    median = Q[4],
    Q075 = Q[5],
    Q095 = Q[6],
    max  = Q[7]
  )
}
```

Let's test the function

**column name**

```{r}
getStats(x, "Sepal.Length") %>%
  Show()
```

**column index**

```{r}
getStats(x, 1) %>%
  Show()
```

## Apply to all groups

```{r}
groupResults <- do.call("rbind", lapply(splits, 
                                        function(x)
                                        {
                                          getStats(x, "Sepal.Length")
                                        }
                                       )
                       )
nrow(groupResults)
```

```{r}
groupResults %>% as_tibble() %>% Show()
```

# Split-Apply-Combine - tidyverse / base hybrid

This hybrid approach uses `split` to create a list of data.frames, and then uses `map_dfr` to map a function over each data.frame and then combine (by rows) into a single data.frame.

```{r}
resultTbl1 <- 
  iris              %>% 
  split(.$Species)  %>%
  purrr::map_dfr(getStats, "Sepal.Length")

resultTbl1 %>% Show()
```

# Split-Apply-Combine - Curent era tidyverse

Need new `getStats` function using `magrittr` pipe

Use new "curly-curly" operator

See [recent June 28 Tweet](https://twitter.com/_lionelhenry/status/1144506303932489728) by @_lionelhenry:

> rlang 0.4.0 is out! Meet curly-curly, a new operator to make it easier to create #rstats functions around #tidyverse pipelines. [Blog post at](https://www.tidyverse.org/articles/2019/06/rlang-0-4-0/)  

`xValue` here cannot be defined by indexing `x`, rather the specified column must be pulled in a tidy way.

```{r}
getStatsTidy <- function(x, columnName)
{
  xValue <- {{x}}   %>% 
            pull({{columnName}})
  
  Q <- quantile(xValue, 
                c(0.00, 0.05, 0.25, 0.50,
                  0.75, 0.95, 1.00),
                na.rm = TRUE)
  
  tibble(
    n     = nrow(x),
    
    mean  = mean(xValue, na.rm = TRUE),
    stdev = sd(xValue,   na.rm = TRUE),
    
    min  = Q[1],
    Q005 = Q[2],
    Q025 = Q[3],
    median = Q[4],
    Q075 = Q[5],
    Q095 = Q[6],
    max  = Q[7]
  )
}
```

```{r}
resultTbl2 <- 
  iris               %>% 
  group_by(Species)  %>%
  nest()             %>%
  mutate( data = purrr::map(data, getStatsTidy, "Sepal.Length"))  %>%
  unnest()

resultTbl2 %>% Show()
```

```{r, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`                                     
